WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-07-08 ms.tex
Add P-P plots

Adding to the challenge of detecting a faint, short-lived optical transient, there is an extreme mismatch between the sky localization accuracy of \ac{GW} detector networks, $\sim 10--500$~deg$^2$ \citep{FairhurstTriangulation,WenLocalizationAdvancedLIGO,FairhurstLocalizationAdvancedLIGO,2011PhRvD..84j4020V,RodriguezBasicParameterEstimation,NissankeLocalization,NissankeKasliwalEMCounterparts,KasliwalTwoDetectors,Grover:2013,SiderySkyLocalizationComparison,FirstTwoYears}, and the \acp{FOV} of 1\nobreakdashes--8\,m\nobreakdashes-class optical telescopes. Wide\nobreakdashes-field optical transient facilites such as BlackGEM (0.6\,m/2.7\,deg$^2$), the \acl{ZTF} \cite{ZTF} (1.2\,m/47\,deg$^2$), DECam (4\,m/3\,deg$^2$), or the \acf{LSST} (8.4\,m/9.6\,deg$^2$), operated in ``target of opportunity'' mode, may be able to tile these large areas rapidly enough to find the one needle in the haystack that is connected to the \ac{GW} event. However, prompt and accurate \ac{GW} position reconstructions will be of the utmost importance for guiding the selection of fields to observe.

@@

\subsection{Properties}

@@

The simulated waveforms were deposited in Gaussian noise that has been filtered to have the \acp{PSD} consistent with \cite{LIGOObservingScenarios}. They were detected using the real\nobreakdashes-time matched\nobreakdashes-filter pipeline, GSTLAL\_INSPIRAL \citep{Cannon:2011vi}. Candidates with estimated \acp{FAR} less than $10^{-2}$\,yr$^{-1}$ were considered to be detections. Because using Gaussian noise results in lower \acp{FAR} than would be calculated in realistically glitchy detector noise, we imposed an additional detection threshold on the network \ac{SNR}, $\hat{\rho} \geq 12$, which has been found to correspond to a comparable \ac{FAR} in the initial \ac{LIGO} runs. For further analysis of the effect of glitchy noise on detection and sky localization, see \cite{BerryLocalization}.

@@

@@

@@

A cumulative histogram of $p$ should be diagonal, within a binomial confidence band. It is already well established that LALINFERENCE\_NEST/MCMC/BAMBI satisfy the $P$\nobreakdashes--$P$ plot test when deployed with accurate templates and reasonable priors. We found at first that \ac{BAYESTAR}'s $P$\nobreakdashes--$P$ plots tended to sag below the diagonal, indicating that though the accuracy (i.e., searched area) was comparable to the stochastic samplers, the \emph{precision} was overstated with confidence intervals that were only about 70\% of the correct area. This was rectified by pre\nobreakdashes-scaling the \acp{SNR} from GSTLAL\_INSPIRAL by a factor of 0.83 prior to running \ac{BAYESTAR}. This correction factor suggests that, for example, an \ac{SNR} 10 trigger from GSTLAL\_INSPIRAL has the \emph{effective information content} of an \ac{SNR} 8.3 signal. The missing information may be due to losses from the discreteness of the template bank, from the \ac{SVD}, from mismatch between the matched\nobreakdashes-filter templates and the simulated signals, from the small but nonzero correlations between masses and intrinsic parameters, or from elsewhere within the detection pipeline. The correction is hard\nobreakdashes-coded into the rapid localization. With it, the $P$\nobreakdashes--$P$ plots are diagonalized without negatively affecting the searched area (see Figure~\ref{fig:pp}).

\begin{figure}
    \begin{center}
        \includegraphics{pp}
    \end{center}
    \caption[\acs{BAYESTAR} $P$\nobreakdashes--$P$ plots]{\label{fig:pp}$P$\nobreakdashes--$P$ plots for \ac{BAYESTAR} localizations in the 2015 and 2016 configurations. The gray lozenge around the diagonal is a target 95\% confidence band derived from a binomial distribution.}
\end{figure}

@@

Information is additive, so the Fisher matrix for the whole detector network is
%
\begin{equation}
    \mathcal{I}_\mathrm{net} = \left(
    \begin{array}{ccccc}
    {\rho_1}^2 A & 0 & \cdots & 0 & {\rho_1}^2 B \\
    0 & {\rho_2}^2 A & & \vdots & {\rho_1}^2 B \\
    \vdots & & \ddots & 0 & \vdots \\
    0 & 0 & \cdots & {\rho_N}^2 A & {\rho_N}^2 B \\
    {\rho_1}^2 B\transpose & {\rho_2}^2 B\transpose & \cdots & {\rho_N}^2 B\transpose & {\rho_\mathrm{net}}^2 C
    \end{array}
    \right).
\end{equation}
%
Now we introduce the change of variables that sacrifices the $N$th detector's extrinsic parameters for the network averages,
%
\begin{align}
    &\begin{array}{c@{\quad\rightarrow\quad}l}
    \varsigma_N & \overline{\varsigma} = \left(\sum_i {\rho_i}^2 \varsigma_i\right) / {\rho_\mathrm{net}}^2, \\
    \gamma_N & \overline{\gamma} = \left(\sum_i {\rho_i}^2 \gamma_i\right) / {\rho_\mathrm{net}}^2, \\
    \tau_N & \overline{\tau} = \left(\sum_i {\rho_i}^2 \tau_i\right) / {\rho_\mathrm{net}}^2, \\
    \end{array}
%
\intertext{and replaces the first $N-1$ detectors' extrinsic parameters with differences,}
%
    &\left.
    \begin{array}{c@{\quad\rightarrow\quad}l}
    \varsigma_i & \delta\varsigma_i = \varsigma_i - \overline{\varsigma} \\
    \gamma_i & \delta\gamma_i = \gamma_i - \overline{\gamma} \\
    \tau_i & \delta\tau_i = \tau_i - \overline{\tau} \\
    \end{array}
    \right\} \text{ for } i = 1, \dots, N - 1.
\end{align}
%
The Jacobian matrix that describes this change of variables is
%
\begin{equation}
    J = \left(
    \begin{array}{cccccc}
    1 & 0 & \cdots & 0 & 1 & 0 \\
    0 & 1 & & 0 & 1 & 0 \\
    \vdots & & \ddots & & \vdots & \vdots \\
    0 & 0 & \cdots & 1 & 1 & 0 \\
    \frac{-{\rho_1}^2}{{\rho_N}^2} & \frac{-{\rho_2}^2}{{\rho_N}^2} & \cdots & \frac{-{\rho_{N-1}}^2}{{\rho_N}^2} & 1 & 0 \\
    0 & 0 & \cdots & 0 & 0 & 1 \\
    \end{array}
    \right).
\end{equation}
%
The transformed network Fisher matrix is block diagonal,
%
\begin{widetext}
\begin{equation}
    \mathcal{I}_\mathrm{net} \rightarrow J\transpose \mathcal{I}_\mathrm{net} J = \left(
    \begin{array}{cccccc}
    {\rho_1}^2(1+\frac{1}{{\rho_1}^4}) A & \frac{{\rho_1}^2 {\rho_2}^2}{{\rho_N}^2} A & \cdots & \frac{{\rho_1}^2 {\rho_{N-1}}^2}{{\rho_N}^2} A & 0 & 0 \\
    \frac{{\rho_1}^2 {\rho_2}^2}{{\rho_N}^2} A & {\rho_2}^2(1+\frac{1}{{\rho_1}^4}) A & & \frac{{\rho_2}^2 {\rho_{N-1}}^2}{{\rho_N}^2} A & 0 & 0 \\
    \vdots & & \ddots & \vdots & \vdots & \vdots \\
    \frac{{\rho_1}^2 {\rho_{N-1}}^2}{{\rho_N}^2} A & \frac{{\rho_2}^2 {\rho_{N-1}}^2}{{\rho_N}^2} A & \cdots & {\rho_{N-1}}^2(1+\frac{1}{{\rho_1}^4}) A & 0 & 0 \\
    0 & 0 & \cdots & 0 & {\rho_\mathrm{net}}^2 A & {\rho_\mathrm{net}}^2 B \\
    0 & 0 & \cdots & 0 & {\rho_\mathrm{net}}^2 B\transpose & {\rho_\mathrm{net}}^2 C \\
    \end{array}
    \right).
\end{equation}
\end{widetext}
%
The top\nobreakdashes-left block contains $N-1$ relative amplitudes, phases, and times on arrival, all potentially correlated with each other. The bottom\nobreakdashes-right block contains the average amplitudes, phases, and times, as well as the masses. The averages and the masses are correlated with each other, but are not correlated with the differences. Because only the differences are informative for sky localization, we drop the intrinsic parameters from the rest of the Fisher matrix calculations in this chapter.

>>>>

Adding to the challenge of detecting a faint, short-lived optical transient, there is an extreme mismatch between the sky localization accuracy of \ac{GW} detector networks, $\sim$10\nobreakdashes--500\,deg$^2$ \citep{FairhurstTriangulation,WenLocalizationAdvancedLIGO,FairhurstLocalizationAdvancedLIGO,2011PhRvD..84j4020V,RodriguezBasicParameterEstimation,NissankeLocalization,NissankeKasliwalEMCounterparts,KasliwalTwoDetectors,Grover:2013,SiderySkyLocalizationComparison,FirstTwoYears}, and the \acp{FOV} of 1\nobreakdashes--8\,m\nobreakdashes-class optical telescopes. Wide\nobreakdashes-field optical transient facilites such as BlackGEM~(0.6\,m/2.7\,deg$^2$), the \acl{ZTF}~(\acsu{ZTF}; 1.2\,m/47\,deg$^2$) \cite{ZTF}, the \acl{DECAM}~(\acsu{DECAM}; 4\,m/3\,deg$^2$), or the \acl{LSST}~(\acsu{LSST}; 8.4\,m/9.6\,deg$^2$), operated in ``target of opportunity'' mode, may be able to tile these large areas rapidly enough to find the one needle in the haystack that is connected to the \ac{GW} event. However, prompt and accurate \ac{GW} position reconstructions will be of the utmost importance for guiding the selection of fields to observe.

@@

\subsection{Properties}
\label{sec:properties}

@@

\subsection{Detection and localization}

The simulated waveforms were deposited in Gaussian noise that has been filtered to have the \acp{PSD} consistent with \cite{LIGOObservingScenarios}. They were detected using the real\nobreakdashes-time matched\nobreakdashes-filter pipeline, GSTLAL\_INSPIRAL \citep{Cannon:2011vi}. Candidates with estimated \acp{FAR} less than $10^{-2}$\,yr$^{-1}$ were considered to be ``detections.'' Because using Gaussian noise results in lower \acp{FAR} than would be calculated in realistically glitchy detector noise, we imposed an additional detection threshold on the network \ac{SNR}, $\hat{\rho} \geq 12$, which has been found to correspond to a comparable \ac{FAR} in the initial \ac{LIGO} runs.\footnote{See \cite{BerryLocalization} for an analysis of the effect of glitchy noise on detection and parameter estimation.} Localizations for the ``detections'' were generated with \ac{BAYESTAR} as well as the functionally equivalent and interchangeable LALINFERENCE\_MCMC, LALINFERENCE\_NEST, and LALINFERENCE\_BAMBI samplers (collectively referred to as LALINFERENCE).

@@

We measured sky localization areas for each event as follows. First, we ranked the \ac{HEALPix} pixels by descending posterior probability. Then, we computed the cumulative sum of the pixels in that order. Finally, we searched for the index of the pixel whose cumulative sum was equal to a given value: for example, 0.9 if we are interested in the 90\% credible area. That pixel index times the area per pixel is the area of the smallest region of the specified credible level.

We can construct a second measure, called the searched area, as the smallest such constructed area that contains the true location of the source. A telescope with a \ac{FOV} that is small compared to the characteristic scale of the posterior would intercept the true location of the source after covering the searched area. This measure is mainly useful because it measures the accuracy of the sky localization independently of the precision: that is to say, it treats the sky map as merely a ranking statistic.

Histograms of the 90\% credible area and the searched area are shown in Figure~\ref{fig:area-hist}, broken down by observing scenario (2015 or 2016) and detector network (HL, HV, LV, or HLV). Note that there are no stastically significant differences in areas between \ac{BAYESTAR} and LALINFERENCE, with the exception in the 2016/HLV configuration, for which some LALINFERENCE sky maps span about an order of magnitude less area than \ac{BAYESTAR}. If we consider only events for which all three detectors contained a signal that was loud enough to trigger the matched\nobreakdashes-filter pipeline, the difference becomes much smaller and insignificant within 95\% error bars. This is because if the signal is too weak to trigger the detection pipeline in one of the detectors, then \ac{BAYESTAR} receives no information about that detector. This issue does not occur in the two\nobreakdashes-detector configurations (HL, HV, or LV) because two or more triggers are required to report a detection candidate.

@@

This is a significant issue for the 2016 configuration, because the most accurate localizations are possible when all three detectors are operating. However, there may be a simple remedy. As we noted in Section~\ref{sec:properties}, the \ac{BAYESTAR} likelihood can be modified to use, instead of the times, phases, and amplitudes on arrival, the full complex matched\nobreakdashes-filter time series from all detectors. The detection pipeline, GSTLAL\_INSPIRAL, would have to be modifed to save and transmit a small interval of the complex \ac{SNR} time series (perhaps a few tens of milliseconds) around the time of each detection candidate. In addition to supplying the missing information for sub\nobreakdashes-threshold signals, this would make \ac{BAYESTAR} mathematically equivalent to the LALINFERENCE analysis, but with the intrinsic parameters fixed to their maximum\nobreakdashes-likelihood values. This idea will be pursued in future work.

@@

A cumulative histogram of $p$ should be diagonal, within a binomial confidence band. It is already well established that LALINFERENCE localizations satisfy the $P$\nobreakdashes--$P$ plot test when deployed with accurate templates and reasonable priors. We found at first that \ac{BAYESTAR}'s $P$\nobreakdashes--$P$ plots tended to sag below the diagonal, indicating that though the accuracy (i.e., searched area) was comparable to the stochastic samplers, the \emph{precision} was overstated with confidence intervals that were only about 70\% of the correct area. This was rectified by pre\nobreakdashes-scaling the \acp{SNR} from GSTLAL\_INSPIRAL by a factor of 0.83 prior to running \ac{BAYESTAR}. This correction factor suggests that, for example, an \ac{SNR} 10 trigger from GSTLAL\_INSPIRAL has the \emph{effective information content} of an \ac{SNR} 8.3 signal. The missing information may be due to losses from the discreteness of the template bank, from the \ac{SVD}, from mismatch between the matched\nobreakdashes-filter templates and the simulated signals, from the small but nonzero correlations between masses and intrinsic parameters, or from elsewhere within the detection pipeline. The correction is hard\nobreakdashes-coded into the rapid localization. With it, the $P$\nobreakdashes--$P$ plots are diagonalized without negatively affecting the searched area (see Figure~\ref{fig:pp}).

\begin{figure*}
    \begin{center}
        \includegraphics[width=0.45\textwidth]{2015-pp}
        \includegraphics[width=0.45\textwidth]{2016-pp}
    \end{center}
    \caption[\acs{BAYESTAR} $P$\nobreakdashes--$P$ plots]{\label{fig:pp}$P$\nobreakdashes--$P$ plots for \ac{BAYESTAR} and LALINFERENCE localizations in the 2015 and 2016 configurations. The gray lozenge around the diagonal is a target 95\% confidence band derived from a binomial distribution.}
\end{figure*}

@@

Information is additive, so the Fisher matrix for the whole detector network is
%
\begin{equation}
    \mathcal{I}_\mathrm{net} = \left(
    \begin{array}{ccccc}
    {\rho_1}^2 A & 0 & \cdots & 0 & {\rho_1}^2 B \\
    0 & {\rho_2}^2 A & & \vdots & {\rho_1}^2 B \\
    \vdots & & \ddots & 0 & \vdots \\
    0 & 0 & \cdots & {\rho_N}^2 A & {\rho_N}^2 B \\
    {\rho_1}^2 B\transpose & {\rho_2}^2 B\transpose & \cdots & {\rho_N}^2 B\transpose & {\rho_\mathrm{net}}^2 C
    \end{array}
    \right).
\end{equation}
%
Now we introduce the change of variables that sacrifices the $N$th detector's extrinsic parameters for the network averages,
%
\begin{align}
    &\begin{array}{c@{\quad\rightarrow\quad}l}
    \varsigma_N & \overline{\varsigma} = \left(\sum_i {\rho_i}^2 \varsigma_i\right) / {\rho_\mathrm{net}}^2, \\
    \gamma_N & \overline{\gamma} = \left(\sum_i {\rho_i}^2 \gamma_i\right) / {\rho_\mathrm{net}}^2, \\
    \tau_N & \overline{\tau} = \left(\sum_i {\rho_i}^2 \tau_i\right) / {\rho_\mathrm{net}}^2, \\
    \end{array}
%
\intertext{and replaces the first $N-1$ detectors' extrinsic parameters with differences,}
%
    &\left.
    \begin{array}{c@{\quad\rightarrow\quad}l}
    \varsigma_i & \delta\varsigma_i = \varsigma_i - \overline{\varsigma} \\
    \gamma_i & \delta\gamma_i = \gamma_i - \overline{\gamma} \\
    \tau_i & \delta\tau_i = \tau_i - \overline{\tau} \\
    \end{array}
    \right\} \text{ for } i = 1, \dots, N - 1.
\end{align}
%
The Jacobian matrix that describes this change of variables is
%
\begin{equation}
    J = \left(
    \begin{array}{cccccc}
    1 & 0 & \cdots & 0 & 1 & 0 \\
    0 & 1 & & 0 & 1 & 0 \\
    \vdots & & \ddots & & \vdots & \vdots \\
    0 & 0 & \cdots & 1 & 1 & 0 \\
    \frac{-{\rho_1}^2}{{\rho_N}^2} & \frac{-{\rho_2}^2}{{\rho_N}^2} & \cdots & \frac{-{\rho_{N-1}}^2}{{\rho_N}^2} & 1 & 0 \\
    0 & 0 & \cdots & 0 & 0 & 1 \\
    \end{array}
    \right).
\end{equation}
%
The transformed network Fisher matrix is block diagonal,
%
\begin{widetext}
\begin{equation}
    \mathcal{I}_\mathrm{net} \rightarrow J\transpose \mathcal{I}_\mathrm{net} J = \left(
    \begin{array}{cccccc}
    {\rho_1}^2(1+\frac{1}{{\rho_1}^4}) A & \frac{{\rho_1}^2 {\rho_2}^2}{{\rho_N}^2} A & \cdots & \frac{{\rho_1}^2 {\rho_{N-1}}^2}{{\rho_N}^2} A & 0 & 0 \\
    \frac{{\rho_1}^2 {\rho_2}^2}{{\rho_N}^2} A & {\rho_2}^2(1+\frac{1}{{\rho_1}^4}) A & & \frac{{\rho_2}^2 {\rho_{N-1}}^2}{{\rho_N}^2} A & 0 & 0 \\
    \vdots & & \ddots & \vdots & \vdots & \vdots \\
    \frac{{\rho_1}^2 {\rho_{N-1}}^2}{{\rho_N}^2} A & \frac{{\rho_2}^2 {\rho_{N-1}}^2}{{\rho_N}^2} A & \cdots & {\rho_{N-1}}^2(1+\frac{1}{{\rho_1}^4}) A & 0 & 0 \\
    0 & 0 & \cdots & 0 & {\rho_\mathrm{net}}^2 A & {\rho_\mathrm{net}}^2 B \\
    0 & 0 & \cdots & 0 & {\rho_\mathrm{net}}^2 B\transpose & {\rho_\mathrm{net}}^2 C \\
    \end{array}
    \right).
\end{equation}
\end{widetext}
%
The top\nobreakdashes-left block contains $N-1$ relative amplitudes, phases, and times on arrival, all potentially correlated with each other. The bottom\nobreakdashes-right block contains the average amplitudes, phases, and times, as well as the masses. The averages and the masses are correlated with each other, but are not correlated with the differences. Because only the differences are informative for sky localization, we drop the intrinsic parameters from the rest of the Fisher matrix calculations in this appendix.
